Modelling the Retinotectal projection - can competition provide a stopping mechanism?

### 2021-08-06 paper/paper.tex
WIP
WIP. Thinking about competition and axon interaction distances

% Define title, empty date and authors
\title {
  Competition can account for stopping in a gradient-following model
  of the retinotectal projection \\
  \emph{or} \\
  Competition provides stopping for the chemoaffinity
  theory which is robust to noise \\
  \emph{or} \\
  Self-organisation of topographically ordered axon connections can take place
  in a noisy environment if axons compete \\
  Time to stop; modelling chemoaffinity with competition \\
}
\date{} % No date please
\author{\Authors}

@@

In our model, space-based competition was omitted and chemoaffinity and
axon-axon interactions were remodelled so that all interactions were via
signal transmission between receptors and ligands, which are expressed in
graded patterns on both retina and tectum. Signals induce attractive or
repulsive interactions; for chemoaffinity the interaction is to ascend or
descend the gradient of ligand expression on the tectum; for axon-axon
interactions, the strength of short-range repulsion (or attraction) is
governed by the receptor-ligand expression on pairs of axons.

@@

Figure~\ref{f:ex} shows colour maps of receptor and ligand
expression, in this case tectal ligand expression maps are linear. Note that
the expression is discretized, and in simulations, the gradient is computed
numerically from the discretized expression. This makes it straightforward to
perform tectal and retinal `graft' manipulations.

@@

At each timestep in a simulation, the position, $\mathbf{x}_{b,t}$, of each
branch, $b$, is updated according to
%
\begin{equation}
\mathbf{x}_{b,t+1} = \mathbf{x}_{b,t} + \mathbf{M}_{b}
\end{equation}
%
where $\mathbf{M}_{b}$, the movement vector, is the sum of a chemoaffinity
effect, $\mathbf{G}_b$, an axon-axon interaction, $\mathbf{I}_b$ and a `border effect',
$\mathbf{B}_b$, which acts to retain branches within the tectal region:
%
\begin{equation}
\mathbf{M}_{b} = m_1 \mathbf{G}_b + m_2 \mathbf{I}_b + m_3 \mathbf{B}_b
\end{equation}
%
evaluated at time $t$. $m_1$, $m_2$ and $m_3$ are scalar parameters.

@@

@@

@@

@@

>>>>

% Define title, empty date and authors
\title {
  Competition can account for stopping in a gradient-following model
  of the retinotectal projection \\
  \emph{or} \\
  Competition provides stopping for the chemoaffinity
  theory which is robust to noise \\
  \emph{or} \\
  Self-organisation of topographically ordered axon connections can take place
  in a noisy environment if axons compete \\
  \emph{or} \\
  Time to stop; modelling chemoaffinity with competition \\
  \emph{or} \\
  Can the retinotectal projection be modelled exclusively by receptor-ligand interactions?
}
\date{} % No date please
\author{\Authors}

@@

In our model, space-based competition was retained (though sometimes omitted)
and chemoaffinity and axon-axon interactions were remodelled so that all
interactions occurred via signal transmission between receptors and ligands,
both of which are expressed in graded patterns on both retina and
tectum. Signals induce attractive or repulsive interactions; for chemoaffinity
the interaction is to ascend or descend the gradient of ligand expression on
the tectum; for axon-axon interactions, the strength of short-range repulsion
(or attraction) is governed by the receptor-ligand expression on pairs of
axons.

@@

Figure~\ref{f:ex} shows colour maps of receptor and ligand expression, in this
case tectal ligand expression maps are linear. Note that the expression is
discretized, and in simulations, the gradient is computed numerically from the
discretized expression. This makes it straightforward to perform tectal and
retinal `graft' manipulations. Coordinates within the simulated retina and
tectum are normalised, with 1 corresponding to 2 mm on the tectum and on the
retina for mouse \citep{reber_relative_2004}. This means that growth cones of
approximate radius 5 $\mu$m = 0.005 mm \citep{goodhill_molecular_2016} have a
radius $r$ = 0.0025 in the arbitrary units of the simulation. For a realistic
simulation based on mouse, the approximate RGC density in the retina is around
44000 cells per retina \citep{jeon_major_1998}. This sets the side length of
the retina to $\sqrt{44000} \approx 200$. This results in a simulation of
prodigious computational size, which although tractable (10 s per step,
requiring several hours, rather than days to run) is not convenient. To
maintain the relationship between growth cone size and cell density, we permit
a smaller number of growth cones ($400\times8=3200$) but keep the fractional
area of the branches at 0.86 ($44000 \times \pi r^2$) (if there is one cone
per axon) or 6.9 (if there are 8 cones per axon).

@@

At each timestep in a simulation, the position, $\mathbf{x}_{b,t}$, at time
$t$ of each branch $b$ is updated according to
%
\begin{equation}
\mathbf{x}_{b,t+1} = \mathbf{x}_{b,t} + \mathbf{M}_{b}
\end{equation}
%
where $\mathbf{M}_{b}$, the movement vector, is the sum of a chemoaffinity
effect, $\mathbf{G}_b$, axon-axon competition, $\mathbf{C}_b$, an axon-axon
receptor-ligand interaction, $\mathbf{I}_b$ and a `border effect',
$\mathbf{B}_b$, which acts to retain branches within the tectal region:
%
\begin{equation}
\mathbf{M}_{b} = m_1 \mathbf{G}_b + m_2 \mathbf{C}_b + m_3 \mathbf{I}_b + m_4 \mathbf{B}_b
\end{equation}
%
evaluated at time $t$. $m_1$--$m_4$ are scalar parameters.

@@

\subsection*{Axon-axon competition}

Competition, in the context of this model, is a repulsive effect which acts
between any two branches, whether they originate from different retinal
ganglion cells or from the same cell. It assumes that there is some kind of
self-other signal (\textbf{be sure to discuss possible origins}) which acts to
discourage two cells from approaching each other.
%
\begin{equation}
\mathbf{C}_b = \frac{1}{|B_b|} \sum_k \hat{\mathbf{x}}_{kb} W
\end{equation}
where the distance based weight $W$ is given by
\begin{equation}
W(d_{kb}) = \begin{cases}
      1 - \frac{d_{kb}}{2r}   & d_{kb} \leq 2r \\
     0 & d_{kb} > 2r
     \end{cases}
\end{equation}
for two growth cones of radius $r$ separated by a distance $d_{kb}$. $B_b$ is
the set of growth cones within a distance $2r$ of branch $b$.

@@

\begin{figure}
\includegraphics[width=\linewidth]{./images/j4_ee_brown.png}
\caption{Brown et al simulation. This is EphA3 knockin - it adds a constant to
half of the RGC's $r_0$ (randomly selected). Compare with S\&G Fig.\,6.}
\label{f:brown}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\linewidth]{./images/j4_ee_reber.png}
\caption{Reber et al simulation. Follow on to Brown et al (same
group). Combines the selective EphA3 knockin with a uniform EphA4 knockdown.}
\label{f:reber}
\end{figure}

@@

\textbf{FIXME: Find the expression functions for the tectum which give a
decent WT result. Perhaps I already did this? Experiment with these and
axon-axon with smaller radius of interaction.}
